HO unification from object language to meta language

### 2024-03-26 paper/main.tex
w

@@

@@

We represent a logic program \emph{run} in \Fo as
a list \emph{steps} $p$ of length $\mathcal{N}$. Each made of a
unification problem between terms $\mathcal{S}_{p_l}$ and
$\mathcal{S}_{p_r}$ taken from the set of all terms $\mathcal{S}$.
The composition of these steps starting from the
empty substitution $\rho_0$ produces the final
substitution $\rho_\mathcal{N}$.
\footnote{If the same rule is used multiple time in a run we
just consider as many copies as needed of the terms composing the
rules, with fresh unification variables each time}
The initial here $\rho_0$ is the empty substitution
%
\newcommand{\progress}{\ensuremath{\mathrm{progress}}\xspace}
\newcommand{\fstep}{\ensuremath{\mathrm{fstep}}\xspace}
\newcommand{\hstep}{\ensuremath{\mathrm{hstep}}\xspace}
\newcommand{\frun}{\ensuremath{\mathrm{frun}}\xspace}
\newcommand{\hrun}{\ensuremath{\mathrm{hrun}}\xspace}
\newcommand{\stepF}[4]{\ensuremath{\fstep(#1,#2,#3) \mapsto #4}}
\newcommand{\stepFD}[5]{%
\ensuremath{#3 #1_{#2_l} \Uo #3 #1_{#2_r} \mapsto #4 \land #5 = #3 \cup #4}}
\newcommand{\stepH}[6]{\ensuremath{\hstep(#1,#2,#3,#4) \mapsto (#5, #6)}}
\newcommand{\stepHD}[6]{\ensuremath{%
#3 #1_{#2_l} \Ue #3 #1_{#2_r} \mapsto #4 \land \progress(#6,#3 \cup #4) \mapsto (#6',#5)}}
\newcommand{\runF}[3]{\ensuremath{\frun(#1,#2) \mapsto #3_{#2}}}
\newcommand{\runFD}[2]{\ensuremath{%
\bigwedge_{p = 1}^{#2} \stepF{#1}{p}{\rho_{p-1}}{\rho_{p}}}}
\newcommand{\runH}[3]{\ensuremath{\hrun(#1,#2) \mapsto #3_{#2}}}
\newcommand{\runHD}[3]{\ensuremath{%
\bigwedge_{p = 1}^{#2} \stepH{#1}{p}{\sigma_{p-1}}{#3_{p-1}}{\sigma_{p}}{#3_p}}}
\newcommand{\deff}{\ensuremath{\stackrel{{\scriptscriptstyle def}}{=\!=\!\!\!=}}}
\def\linkStore{\ensuremath{\mathbb{L}}}

@@

\newtheorem{invariant}{Invariant}

@@

\begin{invariant}[Link left hand side]\ref{inv:linklhs}
  The left hand side of a new link
  is a variable.
\end{invariant}

@@

\subsection{Execution}

@@

\noindent
Note tha the notation \elpiIn{((A ~\Ue~B) C D)} is syntactic sugar for
\elpiIn{((~\Ue~) A B C D)}.

@@

Even though the unification process explained in the previous sections is able
to solve a large number of unification problems, it remains still incomplete:
\wellb is only a subset of terms in \Ho. In order to capture all the unification
properties of \Eo, we need ad-hoc compilation strategies over those subterms
that have been defined as ``problematic''.

\subsection{Compilation}
\subsection{Progress}

@@

and aux

\otext{The following goal necessita v1 (lo scope è usato):\\ \elpiIn{X = lam x\ lam y\ Y y x, X = lam x\ f}}
\otext{The snd unif pb, we have to unif \elpiIn{lam x\ lam y\ Y x y} with \elpiIn{lam x\ f}}
\otext{It is not doable, with the same elpi var}

Invarianti:
A destra della eta abbiamo sempre un termine che comincia per $\lambda x. bla$

\begin{textcode}
  La deduplicate eta:
  - viene chiamata che della forma [variable] -> [eta1] e [variable] -> [eta2]
    (a destra non c'è mai un termine con testa rigida)
  - i due termini a dx vengono unificati con la unif e uno dei due link viene buttato
    NOTA!! A dx abbiamo sempre un termine della forma lam x.VAR x!!!
    Altrimenti il link sarebbe stato risolto!!
  - dopo l'unificazione rimane un link [variabile] -> [etaX]
  - nella progress-eta, se a sx abbiamo una constante o un'app, allora eta-espandiamo
    di uno per poter unificare con il termine di dx.
\end{textcode}

@@

>>>>

\newtheorem{invariant}{Invariant}
\crefname{invariant}{invariant}{invariants}

@@

\newcommand{\lhs}{\ensuremath{\mathrm{lhs}}\xspace}
\newcommand{\rhs}{\ensuremath{\mathrm{rhs}}\xspace}

@@

We represent a logic program \emph{run} in \Fo as
a list \emph{steps} $p$ of length $\mathcal{N}$. Each made of a
unification problem between terms $\mathcal{S}_{p_l}$ and
$\mathcal{S}_{p_r}$ taken from the set of all terms $\mathcal{S}$.
The composition of these steps starting from the
empty substitution $\rho_0$ produces the final
substitution $\rho_\mathcal{N}$.
\footnote{If the same rule is used multiple time in a run we
just consider as many copies as needed of the terms composing the
rules, with fresh unification variables each time}
The initial here $\rho_0$ is the empty substitution
%
\newcommand{\progress}{\ensuremath{\mathrm{progress}}\xspace}
\newcommand{\fstep}{\ensuremath{\mathrm{fstep}}\xspace}
\newcommand{\hstep}{\ensuremath{\mathrm{hstep}}\xspace}
\newcommand{\frun}{\ensuremath{\mathrm{frun}}\xspace}
\newcommand{\hrun}{\ensuremath{\mathrm{hrun}}\xspace}
\newcommand{\stepF}[4]{\ensuremath{\fstep(#1,#2,#3) \mapsto #4}}
\newcommand{\stepFD}[5]{%
\ensuremath{#3 #1_{#2_l} \Uo #3 #1_{#2_r} \mapsto #4 \land #5 = #3 \cup #4}}
\newcommand{\stepH}[6]{\ensuremath{\hstep(#1,#2,#3,#4) \mapsto (#5, #6)}}
\newcommand{\stepHD}[6]{\ensuremath{%
#3 #1_{#2_l} \Ue #3 #1_{#2_r} \mapsto #4 \land \progress(#6,#3 \cup #4) \mapsto (#6',#5)}}
\newcommand{\runF}[3]{\ensuremath{\frun(#1,#2) \mapsto #3_{#2}}}
\newcommand{\runFD}[2]{\ensuremath{%
\bigwedge_{p = 1}^{#2} \stepF{#1}{p}{\rho_{p-1}}{\rho_{p}}}}
\newcommand{\runH}[3]{\ensuremath{\hrun(#1,#2) \mapsto #3_{#2}}}
\newcommand{\runHD}[3]{\ensuremath{%
\bigwedge_{p = 1}^{#2} \stepH{#1}{p}{\sigma_{p-1}}{#3_{p-1}}{\sigma_{p}}{#3_p}}}
\newcommand{\deff}{\ensuremath{\stackrel{{\scriptscriptstyle def}}{=\!=\!\!\!=}}}
\newcommand{\linkStore}{\ensuremath{\mathbb{L}}\xspace}

@@

@@

\begin{invariant}[Link left hand side]\label{inv:linklhs}
  The left hand side of a new link
  is a variable.
\end{invariant}

@@

\subsection{Execution}
\label{sec:execution}

@@

\noindent
Note tha the infix notation \elpiIn{((A ~\Ue~B) C D)} is syntactic sugar for
\elpiIn{((~\Ue\!\!\!~) A B C D)}.

@@

$\eta$-reduction is an equivalence relation over terms where a term of the form
$\lambda x.t \appsep x$ can be converted to $t$ any time $x$ does not occur as a
free variable in $t$. We call $\lambda x. t \appsep x$ the $\eta$-expanded
version of $t$. The implementation of the \elpiIn{comp} relation given in
\cref{sec:compilation} compiles the \Fo{} terms $t_1 = $\elpiIn{flam x\ fapp
[fuva A, x]} and $t_2 =$ \elpiIn{fcon "f"} into the \Ho terms $t'_1 =
$\elpiIn{lam x\ uva A' x} and $t'_2 =$ \elpiIn{fcon "f"} with mapping
\mapping{A}{A'}{1}. However, if the oracle sets \elpiIn{A'} to the constant
\elpiIn{"f"}, the unification of $t'_1$ and $t'_2$ in the meta language will
fail even though $t_1 \Eo t_2$. The reason of this failure is attributed to the
fact that $t'_1 =$\elpiIn{lam x\ app[con "f", x]} cannot be unified with $t'_2
=$\elpiIn{con "f"} since the two terms have different rigid heads. We solve this
unification problem by adapting the \elpiIn{comp} relation such that it
recognizes \maybeeta subterms $s$ and replaces them with fresh \Ho variables
$v$. This link between the variable $v$ and the subterm $t$ is stored in what we
call \linketa which is an object with the following type

\elpiIn{type link-eta tm -> tm -> baselink}

\noindent
where, as sketched in \cref{sec:lang-spec}, the term on the left hand side
(\lhs) is linked with its left counterpart (\rhs).

\linketa are added in the link store (\linkStore) and activated when special
conditions are satisfied on \lhs or \rhs. These link activations are managed by
extending the \elpiIn{progress1} predicate (see \cref{sec:execution}).
We claim that \linketa progression does not contradict \cref{inv:linklhs} and we
add the following invariant:

\begin{invariant}[\linketa \rhs]
  The \rhs of a \linketa having the shape $\lambda x.F_x$
  where $F_x$ is a term not starting with the \elpiIn{lam} constructor.
  \label{inv:link-eta-right}
\end{invariant}

In the next three subsections we explain how we detect \maybeeta terms, how we
compile them and how \linketa are activated during the execution of the program
and provide justification for why \cref{inv:linklhs,inv:link-eta-right} remain true.

\subsection{Detection of \maybeeta}

Compiling term with \maybeeta terms forces us to determine if, $\lambda x. T_x$,
for any term $T$ having $x$ in scope, can be a $\eta$-expansion, i.e. under a
given substitution $\sigma$, we have $\sigma (\lambda x.T_x) = t$. This
\maybeeta detection is not a trivial operation as it may seems.
%
\begin{gather}
  \lambda x. f \appsep A_x\\
  \lambda x. f \appsep x \appsep A_x\\
  \lambda x. \lambda y. f \appsep A_x \appsep B_{y x}\\ \label{eq:lamlam}
\end{gather}

In the examples above, the first expression is a \maybeeta since $A_x$ can
reduce to $x$, the second one is not a \maybeeta since for any substitution for
$A_x$, $x$ is not free in $f \appsep x$. The third equation is a bit more
complicated since, we have a spine of lambdas, this means that the whole term
is a \maybeeta, if the inner $\lambda$-term is an $\eta$-expansion of a term
$t$, and $t$ can be reduced to a term on the form $t' x$ where $x$ is not free
in $t'$. Indeed, eq.~\ref{eq:lamlam} is a \maybeeta under the substitution 
$\sigma = \{A \mapsto \lambda x.x, B \mapsto \lambda x. \lambda y. x\}$.

TODO: clarify this
As a remark, note that $\lambda x. f \appsep A_x \appsep x$ is a \maybeeta,
since, despite, $x$ occurs in $f \appsep A_x$, it is still possible that this
subterm does not use $x$, for example if $A$ is a function on the form $\lambda
x.a$, where $a$ is a defined constant. In this case, the \maybeeta should
consider that the bound variable \elpiIn{x} does not ``rigidly'' occur in 
the given subterm.

We can now define more formally the two auxiliary relation we need for \maybeeta
detection:

\begin{definition}[reduce-to]
  A term $t$ reduce to a name $x$, if $\exists \sigma, \sigma t = x$. In
  particular, for any term $t$, $\lambda x_1 \ldots x_n.t_{x_1\ldots x_n}$
  reduces to a bound variable $x$ if one of the three following cases is
  satisfied: 1) $n = 0$ and $t = x$; 2) $t$ is the application of $x$ to a list
  of term $l$ and each $l_i$ reduces to $x_i$; 3) $t$ is a variable with scope
  $s$, and for any $bv \in [x|\vec{x}]$, it exists a $s_i$, such that $s_i$
  reduces to $bv$
\end{definition}

\begin{definition}[occurs-rigidly]
  A name $x$ occurs rigidly in a term $t$, if $\forall \sigma, \exists t' \in
  \subterm{t}, t' = x$, i.e. $x$ does not appear in the scope of any variable
  in $t$.
\end{definition}

Finally, we can derive the implementation for \maybeeta detection:

\begin{definition}[\maybeeta detection]
  For any term $t$, $\lambda x_1 \ldots x_n.t_{x_1\ldots x_n}$ is a \maybeeta if
  1) either $t$ is the application of a term with argument $l_1 \ldots l_m$ such that 
  $m \geq n$, every $l_{m-n+i}$ reduce to $x_i$, and no $l_{1\ldots m-n-1}$
  reduces to a name in $\vec{x}$; 2) or $t$ is a variable with scope $s$ and
  each $x_i$ it exists $s_j$ such that $s_j$ reduces $x_i$.
\end{definition}

As a final remark, the \maybeeta detection defined just before is an
over-approximation, in the sense that there exists some terms $t$ considered as
\maybeeta, such that forall substitution $\sigma$, $\sigma t$ is not an
$\eta$-expansion. A small example is $\lambda x. f \appsep A_x \appsep A_x$. On
the other hand, we also point out that the \maybeeta detection spot out
potential $\eta$-expansion for terms that are not in \llambda. For example,
$\lambda x.F \appsep G_x$ is considered as \maybeeta, since we have the
application of term whose argument can reduce to $x$.

The implementation we propose for the \maybeeta relation is given below.

% Here a complex maybeeta example
% \begin{gather}
%   T = \lambda y. f \appsep A_{xy} \appsep (B \appsep a \appsep (\lambda z.y\appsep C_z)) \appsep D_x
% \end{gather}

Thanks to the \elpiIn{maybe-eta} predicate, we can detect ``$\eta$-problematic''
terms and, consequently replace them with fresh \Ho unification variables at
compilation time. The code below illustrate how this relation is used to for
term compilation.

@@

This rule, to be inserted just before rule~\ref{rule:complam} from the code in
\cref{sec:compilation}, verifies if the \Fo{} term $t$ received in entry is a
\maybeeta. Let $\lambda t'$ be the compiled version of $t$, then the fresh
variable \elpiIn{A} returned as the new \Ho term as in scope all the free names
in $t'$. The critical part of this compilation is the creation of the \linketa,
which links the variable \elpiIn{A} with $t$. This link creation enforce
\cref{inv:linklhs} and \cref{inv:link-eta-right}, since clearly the \lhs is a
variable and the \rhs is a term starting with the \elpiIn{lam} constructor. We
can also note that for any name $x$, $tx = t' \appsep x$ never start with
\elpiIn{lam}, since by contradiction, if 1) $tx$ is an $\eta$-expansion, then
the recursive call to \elpiIn{comp} would have replaced $tx$ with a fresh
variable, which contradict the hypothesis and if 2) $t'$ is not an
$\eta$-expansion, then neither $\lambda t'$ is.

\linketa are meant to suspend the unification of two terms, that is, if we have
a \maybeeta term $t$ which should be unified with a term $t'$ we don't want to
unify them with \Ue. As said before this would break unification but it can also
introduce \maybeeta in the substitution of \Ho.

In order to activate a \linketa, we need to implement new rules for the
\elpiIn{progress1} predicate. There are two cases making a \linketa to progress,
1) \lhs is instantiated to a rigid term, in this case \lhs is unified with \rhs.
TODO: the right eta hand side is eta-expanbded if it is an app/con; 2) \rhs can
be $\eta$-reduced to a term with rigid head, in this case \lhs and \rhs are
again unified. If one of these two condition is satisfied, the link has
fulfilled is task and can be removed from the list of suspended links; if none
the condition succeeds, the link is kept for a further iteration of
\elpiIn{progress}.

TODO: dire che scendiamo sotto i vari abs che formano il contesto? 
TODO: example for case 1: $\lambda x. \lambda y. F \appsep y \appsep x = f$
TODO: example for case 2: $\lambda x. \lambda y. F \appsep y \appsep x = G, F = \lambda x. \lambda y. a$

A second way to progress \linketa, that we call \linketa deduplication, is to
check if \linkStore contains two \linketa $l_1$ and $l_2$ with same \lhs. This
situation occurs if two \maybeeta terms are unified with a same unification
variable. In this case, we can unify the $l_1$ and $l_2$ \rhs (that, by
construction are both on the form $\lambda x. T_x$) and remove one of the two
links.

TODO: example for this: $\lambda x. \lambda y. F \appsep y \appsep x = X,
\lambda x. \lambda y. F \appsep y \appsep x = Y$

We can note that the 1) insertion of these rules for \elpiIn{progress1} do not
prevent the termination of \elpiIn{progress}, since, a link activation runs
terminating operations (such as term unification and link-removal); 2) \linketa
deduplication runs again terminating operations and 3) if none of these two 
situations is performed, then the substitution and \linkStore
Note that if the link remain suspended \elpiIn{progress} continues to terminate,
since, it would mean that the condition of the \elpiIn{if}-branchement succeds
and the \elpiIn{progress} terminates its execution.

\begin{lemma}
  We never add eta-expansions in the substitution
\end{lemma}

\begin{proof}[Proof sketch]
\end{proof}

TODO: we can however have $\lambda x. F_x$ if we know that $F$ does not reduce
to $T x$ where x is not free in $T$.

@@

\paragraph{Detection of \maybebeta} TODO: \dots

\paragraph{Compilation with \linkbeta}
